Statistical analysis of periodic data in neuroscience

### 2020-12-23 manuscript.Rmd
Completed first draft of manuscript

---
title: "Statistical analysis of periodic data in neuroscience"
author: "Daniel H. Baker"
date: "13/12/2020"
output: pdf_document
bibliography: bibliography.bib
biblio-style: apalike
---

@@

library(pals)

@@

@@

## Abstract

## Background

A widely used paradigm in many branches of neuroscience is to drive the nervous system using periodic stimuli. This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations. These periodic responses, often called the _steady-state_ or _frequency following_ response, can be recorded from single neurons [@Enroth-Cugell1966] and local field potentials [@Morrone1987] using invasive methods, or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans [@Norcia2015] and in diverse animal species including insects [@Afsari2014], birds [@Porciatti1990], rodents [@Hwang2019] and primates [@Nakayama1982]. Steady-state methods are used to measure early sensory responses in vision [@Regan1966], hearing [@Rees1986] and somatosensation [@Snyder1992], and closely related paradigms have been developed to target specific stimulus features such as orientation [@Braddick1986], and facial expression [@Gray2020] and identity [@Liu-Shuang2014]. In fMRI research, _travelling wave_ methods [@Engel1994; @Sereno1995] are used to map the retinotopic responses of early visual cortex, again using stimuli that change periodically, this time in spatial position. Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way [@Spitschan2014].

@@

For pointwise and pairwise comparisons, Hotelling's $T^2$ statistic [@Hotelling1931] is a multivariate extension of the t-test. For the one-sample case, the test statistic is defined as:

@@

where _N_ is the number of observations, $\bar{x}$ is the multivariate sample mean, $\mu$ is the point of comparison, $C^{-1}$ is the inverse covariance matrix, and $'$ denotes vector transposition. Conceptually, the $T^2$ statistic extends the univariate t-statistic by incorporating the covariance between the dependent variables. Two-sample and paired variants are also available, and the test can be applied to an arbitrary number of dependent variables (though here we will consider only the bivariate case).

@@

In the present paper, we develop best practice guidelines for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms. We first demonstrate why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed. We then investigate conditions under which we should use either the $T^2$ or $T^2_{circ}$ statistic. The range of sample sizes and effect sizes where $T^2_{circ}$ is more sensitive is identified. We develop a novel method for testing the assumptions of the $T^2_{circ}$ statistic, based on calculating the condition index of a multivariate data set. We then discuss appropriate methods for identifying outliers and calculating error bars. Finally, we demonstrate the proposed techniques on two example SSVEP data sets, and recommend some best practice guidelines for analysis decisions. All data and scripts used to produce this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats). These include _R_ functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index statistic proposed in this paper.

## Fourier amplitudes violate parametric assumptions of univariate statistics

Many studies use parametric t-tests and analysis of variance (ANOVA) to analyse periodic data. Specifically, the amplitude component of the Fourier spectrum at the stimulation frequency is used as the dependent variable, discarding the phase information. This is problematic, because the amplitude is an absolute quantity, and can never fall below zero. Distributions of amplitudes for weak signals are therefore likely to be positively skewed, and in general will violate the assumption of normality.

@@

The upper row of Figure \@ref(fig:amphists) shows scatterplots of simulated Fourier components, expressed using real (x) and imaginary (y) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \@ref(fig:amphists) shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d > 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]). One consequence of this is that t-tests will typically have an inflated Type 1 error (false positive) rate for many signals encountered empirically.

Typical solutions, such as log-transforming the data, are unlikely to be equally applicable to all conditions. For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal. Applying a transform to both conditions is therefore problematic. Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents. Instead, the bivariate statistics discussed in the introduction avoid these issues and have greater statistical power, as we will demonstrate in the next section.

## Conditions under which $T^2_{circ}$ is more sensitive than $T^2$

Using the $T^2$ and $T^2_{circ}$ allow us to retain the phase information, and therefore have greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values. Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$. Here we replicate and extend these simulations (see Figure \@ref(fig:powerfig)), and demonstrate that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f). However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N > 32) and large effect sizes (d > 2 when N > 8)

```{r powerfig, fig.cap=".", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}

@@

par(pty="s")
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,4,1)    # locations of tick marks on y axis
ticklabelsx <- 2^ticklocsx        # set labels for x ticks
ticklabelsy <- seq(0,4,1)    # set labels for y ticks
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(expression(T^2),cex.main=2)
image(slist,dlist,tsqpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
text(1.1,3.8,'(d)',cex=2)
legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white')

@@

## Limitations of $T^2_{circ}$ when assumptions are violated

```{r falsealarms, fig.asp=1.4, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b). Estimates are for 100000 simulations per condition, with N=10 observations. The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}

@@

plotlims <- c(-1,1,0,0.1) 
ticklocsx <- seq(-1,1,0.5)    # locations of tick marks on x axis
ticklocsy <- seq(0,0.1,0.025)    # locations of tick marks on y axis
ticklabelsx <- ticklocsx        # set labels for x ticks
ticklabelsy <- ticklocsy*100    # set labels for y ticks

plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2) 
title(xlab="Correlation (R)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
title(ylab="Percent significant", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)

@@

plotlims <- c(-1,1,0,0.1) 
ticklocsx <- seq(-1,1,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,0.1,0.025)    # locations of tick marks on y axis
ticklabelsx <- c('1/10','1','10')        # set labels for x ticks
ticklabelsy <- ticklocsy*100    # set labels for y ticks

plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2) 
title(xlab="Variance ratio (y/x)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
title(ylab="Percent significant", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)

@@

## A novel method to test the assumptions of $T^2_{circ}$

Despite the severe consequences of violating the assumptions of the $T^2_{circ}$ statistic, there is currently no accepted test for those assumptions that could be applied to real data. Victor and Mast [-@Victor1991] suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single individual, whereas data from multiple participants may be less likely to exhibit independence of the real and imaginary components [@Pei2017].

@@

$pdf = (n-1)2^{n-1}\frac{x^2 - 1}{(x^2 + 1)^n}x^{(n-2)}$

@@

$pdf = (n-2)2^{n-2}\frac{x^2 - 1}{(x^2 + 1)^{(n-1)}}x^{(n-3)}$

In Figure \@ref(fig:distcomparison)a we show an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100000 random data sets for n=10 (blue shading). The vertical lines show the critical (95%) threshold for the analytic and simulated results. Figure \@ref(fig:distcomparison)b shows how these thresholds change as a function of the number of observations, and it is clear that the modified expression (red) most closely approximates the simulation results (blue).

```{r distcomparison, fig.cap='d', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}

@@

plotlims <- c(1,5,0,1) 
ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2) 
title(xlab="Eigenvalue ratio", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
title(ylab="Probability", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)

@@

a <- hist(simvect,breaks = 200, plot=FALSE)
axvals <- a$mids
ayvals <- a$counts/max(a$counts)
polygon(c(1,axvals,1), c(0,ayvals,0), col=addalpha(nbdtpal[3],alpha=0.2),border=NA)
simcrit <- quantile(simvect,0.95)
lines(c(simcrit,simcrit),c(0,1),col=nbdtpal[3],lwd=2,lty=2)

@@

lines(3:25,edelman1,lwd=2,col=nbdtpal[1])
lines(3:25,edelman2,lwd=2,col=nbdtpal[2])
lines(3:25,simratios,lwd=2,col=nbdtpal[3])
legend(12, 25, c("Edelman","Modified","Simulation"), cex=0.8, col=nbdtpal, lty=1, lwd=3, box.lwd=2)

@@

## Methods for calculating error bars for Fourier amplitudes

## Applying these methods to real data sets

### Human visual steady-state data

### Mouse auditory steady-state data

https://www.nature.com/articles/s41597-020-00621-z
https://doi.gin.g-node.org/10.12751/g-node.e5tyek/

## Recommendations for analysing periodic data

Multi/Bivariate statistics should be used for phase-locked Fourier data

Always remove outliers with a Mahalanobis distance > 3

>>>>

---
title: "Statistical analysis of periodic data in neuroscience"
author: "Daniel H. Baker"
date: "13/12/2020"
output:
  bookdown::pdf_document2:
    fig_caption: yes 
    toc: false   
    keep_tex: true
  html_document:
    df_print: paged
  pdf_document: default
bibliography: bibliography.bib
biblio-style: apalike
---

@@

# reasonably compact code to check which packages are installed, 
# install the missing ones, and activate all
packagelist <- c('bookdown','pals','shape','diagram','HDMD')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

anovacirc.test <- function(data){
  # analysis of variance using complex data
  # an extension of the logic of the t-squared circ statistic

  grouplabels <- data$grouplabels
  datavals <- data$simdata
  factorlist <- levels(grouplabels)

  grandmean <- mean(datavals)

  groupmeans <- NULL
  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}

  SSM <- 0
  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}

  dfM <- 2*(nlevels(grouplabels)-1)

  allresiduals <- 1:length(datavals)
  for (n in 1:nlevels(grouplabels)){allresiduals[which(grouplabels==factorlist[n])] <- abs(datavals[which(grouplabels==factorlist[n])] - groupmeans[n])}

  SSR <- sum(allresiduals^2)
  dfR <- 2*(length(datavals)-nlevels(grouplabels))
  MSM <- SSM/dfM
  MSR <- SSR/dfR
  Fratio <- MSM/MSR
  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)

  output <- NULL
  output$Fratio <- Fratio
  output$pvalue <- pvalue
  output$SSM <- SSM
  output$SSR <- SSR
  output$dfM <- dfM
  output$dfR <- dfR
  output$MSM <- MSM
  output$MSR <- MSR

  return(output)
}

anovacircR.test <- function(data){
  # analysis of variance using complex data
  # an extension of the logic of the t-squared circ statistic
  # this version for repeated measures designs

  grouplabels <- data$grouplabels
  factorlist <- levels(grouplabels)
  participantlabels <- data$participant
  participantlist <- levels(participantlabels)
  datavals <- data$simdata

  grandmean <- mean(datavals)

  SSW <- 0
  for (n in 1:nlevels(participantlabels)){SSW <- SSW + sum(abs(datavals[which(participantlabels==participantlist[n])] - mean(datavals[which(participantlabels==participantlist[n])]))^2)}
  dfW <- 2*(nlevels(participantlabels)*(nlevels(grouplabels)-1))

  groupmeans <- NULL
  for (n in 1:nlevels(grouplabels)){groupmeans[n] <- mean(datavals[which(grouplabels==factorlist[n])])}
  SSM <- 0
  for (n in 1:nlevels(grouplabels)){SSM <- SSM + length(which(grouplabels==factorlist[n]))*abs(groupmeans[n]-grandmean)^2}
  dfM <- 2*(nlevels(grouplabels)-1)

SSR <- SSW - SSM
dfR <- dfW - dfM

  MSM <- SSM/dfM
  MSR <- SSR/dfR
  Fratio <- MSM/MSR
  pvalue <- pf(Fratio,df1=dfM,df2=dfR,lower.tail=FALSE)

  output <- NULL
  output$Fratio <- Fratio
  output$pvalue <- pvalue
  output$SSW <- SSW  
  output$SSM <- SSM
  output$SSR <- SSR
  output$dfW <- dfW
  output$dfM <- dfM
  output$dfR <- dfR
  output$MSM <- MSM
  output$MSR <- MSR

  return(output)
}

getel <- function(compdata){
  A <- cov(compdata)
  ctr    <- colMeans(compdata) 
  RR     <- chol(A)                               # Cholesky decomposition
  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
  eigVal  <- eigen(A)$values
  eigVec  <- eigen(A)$vectors
  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
  ellRot  <- eigVec %*% t(ellBase) 
  return(ellRot)
  }

v4Interp <- function(df, xo, yo, rmax = .75, gridRes = 67) {
  ## Create a function to perform Matlab's v4 interpolation.
  ## Takes as input a data-frame with columns x, y, and z (x co-ordinates, y co-ordinates, and amplitude)
  ## and variables xo and yo, the co-ordinates which will be use to create a grid for interpolation
  xo <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
  yo <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
  xy <- df$x + df$y*sqrt(as.complex(-1))
  d <- matrix(rep(xy,length(xy)),nrow = length(xy), ncol = length(xy))
  d <- abs(d - t(d))
  diag(d) <- 1
  g <- (d^2) * (log(d)-1)   # Green's function.
  diag(g) <- 0
  weights <- qr.solve(g,df$z)
  xy <- t(xy)
  outmat <- matrix(nrow = gridRes,ncol = gridRes)
  for (i in 1:gridRes){
    for (j in 1:gridRes) {
      test4 <- abs((xo[i,j] + sqrt(as.complex(-1))*yo[i,j]) - xy)
      g <- (test4^2) * (log(test4)-1)
      outmat[i,j] <- g %*% weights}}
  outDf <- data.frame(x = xo[,1],outmat)
  names(outDf)[1:length(yo[1,])+1] <- yo[1,]
  return(outDf)}

@@

# Abstract

Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli. Neural signals can then be recorded, and include phase-locked oscillations at the stimulation frequency. The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components. However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and weak signals might be missed if the phase information is discarded. An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components. Here we compare the performance of two multivariate extensions of the T-test, Hotelling's $T^2$ and a variant called $T^2_{circ}$. We develop a novel test of the assumptions of $T^2_{circ}$ based on the condition index, and a heuristic for excluding outliers using the Mahanalobis distance. We then extend the statistic to multi-level designs, and propose a new statistical test termed $ANOVA^2_{circ}$. This has identical assumptions to $T^2_{circ}$, and is shown to be more sensitive than MANOVA when these assumptions are met. We demonstrate how these tests might be applied to two publicly available empirical data sets, and suggest concrete guidance for choosing which test to run. Implementations of these novel tools are provided as _R_ functions, and we hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.

# Background

A widely used paradigm in many branches of neuroscience is to drive the nervous system using periodic stimuli. This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations. These periodic responses, often called the _steady-state_ or _frequency following_ response, can be recorded from single neurons [@Enroth-Cugell1966] and local field potentials [@Morrone1987] using invasive methods, or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans [@Norcia2015] and in diverse animal species including insects [@Afsari2014], birds [@Porciatti1990], rodents [@Hwang2019] and primates [@Nakayama1982]. Steady-state methods are used to measure early sensory responses in vision [@Regan1966], hearing [@Rees1986] and somatosensation [@Snyder1992], and closely related paradigms have been developed to target specific stimulus features such as orientation [@Braddick1986], and facial expression [@Gray2020] and identity [@Liu-Shuang2014]. In fMRI research, _travelling wave_ methods [@Engel1994; @Sereno1995] are used to map the retinotopic responses of early visual cortex using stimuli that change periodically in spatial position. Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way [@Spitschan2014].

@@

For pointwise and pairwise comparisons, Hotelling's $T^2$ statistic [@Hotelling1931] is a multivariate extension of the T-test. For the one-sample case, the test statistic is defined as:

@@

where _N_ is the number of observations, $\bar{x}$ is the multivariate sample mean, $\mu$ is the point of comparison, $C^{-1}$ is the inverse covariance matrix, and $'$ denotes vector transposition. Conceptually, the $T^2$ statistic extends the univariate T-statistic by incorporating the covariance between the dependent variables. Two-sample and paired variants are also available, and the test can be applied to an arbitrary number of dependent variables (though here we will consider only the bivariate case).

@@

In the present paper, we develop best practice guidelines for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms. We first demonstrate why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed. We then investigate conditions under which we should use either the $T^2$ or $T^2_{circ}$ statistic. The range of sample sizes and effect sizes where $T^2_{circ}$ is more sensitive is identified. We develop a novel method for testing the assumptions of the $T^2_{circ}$ statistic, based on calculating the condition index of a multivariate data set. Appropriate methods for identifying outliers using the Mahalanobis distance are discussed, and a heuristic proposed. Next the logic of $T^2_{circ}$ is extended to situations with more than two levels of the independent variable, and the performance of this novel $ANOVA^2_{circ}$ statistic is compared to MANOVA. Finally, we demonstrate the proposed techniques on two example SSVEP data sets, and recommend some best practice guidelines for analysis decisions.

All scripts used to generate this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats). These include _R_ functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.

# Fourier amplitudes violate parametric assumptions of univariate statistics

Many empirical studies use univariate T-tests or analysis of variance (ANOVA) to analyse periodic data. Specifically, the amplitude component of the Fourier spectrum at the stimulation frequency is used as the dependent variable, discarding the phase information. This is problematic, because the amplitude is an absolute quantity, and can never fall below zero. Distributions of amplitudes for weak signals are therefore positively skewed, and will generally violate the assumption of normality.

@@

The upper row of Figure \@ref(fig:amphists) shows scatterplots of simulated Fourier components, expressed using real (x) and imaginary (y) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \@ref(fig:amphists) shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d > 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]). One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.

Typical solutions, such as log-transforming the data, are unlikely to be equally applicable to all conditions. For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal. Applying a transform to both conditions is therefore problematic. Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents. Instead, the bivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.

# Conditions under which $T^2_{circ}$ is more sensitive than $T^2$

Using the $T^2$ or $T^2_{circ}$ statistic allows us to retain the phase information, and therefore have greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values. Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$. This involved generating random data sets of different sample sizes, and different signal strengths, and comparing the number of such tests where each statistic was significant. Their simulations show the largest advantage for $T^2_{circ}$ for effect sizes around d = 1 (where the mean signal strength is equal to the standard deviation of the data). The advantage appeared to be stronger for smaller sample sizes.

Here we replicate and extend these simulations (see Figure \@ref(fig:powerfig)), and confirm that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f). However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N > 32) and large effect sizes (d > 2 when N > 8). These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for sample sizes, so its use should be restricted to small sample studies.

```{r powerfig, fig.cap="Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes. Panels a-c replicate conditions reported by Victor and Mast (1991). Panels (d) and (e) show a wider range of conditions for each statistic. Panel (f) shows the difference between the two statistics, with contour lines indicating differences of 0.02, 0.05, 0.1, 0.2 and 0.4.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}

@@

par(pty="s")
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,4,1)    # locations of tick marks on y axis
ticklabelsx <- 2^ticklocsx        # set labels for x ticks
ticklabelsy <- seq(0,4,1)    # set labels for y ticks
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(expression(T^2),cex.main=2)
image(slist,dlist,tsqpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
text(1.1,3.8,'(d)',cex=2)
legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white',title='Proportion')

@@

# Limitations of $T^2_{circ}$ when assumptions are violated

Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on meeting the test's more stringent assumptions. The two variables must be independent, and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \@ref(fig:falsealarms) shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (dashed line). However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the variances of the two dependent variables more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2. In contrast, the $T^2$ statistic, which explicitly takes account of the covariance matrix (see equation \@ref(eq:t2eq)) shows no increase (black curves).

```{r falsealarms, fig.width=6, fig.height=9, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b). Estimates are for 100000 simulations per condition, with N=10 observations. The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}

@@

plotlims <- c(-1,1,0,0.1) 
ticklocsx <- seq(-1,1,0.5)    # locations of tick marks on x axis
ticklocsy <- seq(0,0.1,0.02)    # locations of tick marks on y axis
ticklabelsx <- ticklocsx        # set labels for x ticks
ticklabelsy <- ticklocsy    # set labels for y ticks

plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Correlation (R)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=2, cex.lab=1.5)

@@

plotlims <- c(-1,1,0,0.1) 
ticklocsx <- seq(-1,1,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,0.1,0.02)    # locations of tick marks on y axis
ticklabelsx <- c('1/10','1','10')        # set labels for x ticks
ticklabelsy <- ticklocsy    # set labels for y ticks

plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx)     # add the tick labels
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Variance ratio (y/x)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=2, cex.lab=1.5)

@@

One possible remedy to control the Type I error rate would be to adjust the alpha level to compensate. However, this will reduce the statistical power of the $T^2_{circ}$ test, and its advantage over $T^2$ is relatively marginal in most situations to begin with (see Figure \@ref(fig:powerfig)). What is required is a method to objectively assess whether the assumptions of $T^2_{circ}$ hold, which we develop in the following section.

# A novel method to test the assumptions of $T^2_{circ}$

Despite the severe consequences of violating the assumptions of the $T^2_{circ}$ statistic (see Figure \@ref(fig:falsealarms)), there is currently no accepted test for those assumptions that could be applied to an empirical data set. Victor and Mast [-@Victor1991] suggest that their test should be applicable to multiple repetitions of a stimulus condition collected from a single participant, whereas data pooled across multiple participants may be less likely to exhibit independence of the real and imaginary components [see also @Pei2017].

@@

\begin{equation}
\label{eq:edelman1}
pdf = (n-1)2^{n-1}\frac{x^2 - 1}{(x^2 + 1)^n}x^{(n-2)},
\end{equation}

@@

\begin{equation}
\label{eq:edelman2}
pdf = (n-2)2^{n-2}\frac{x^2 - 1}{(x^2 + 1)^{(n-1)}}x^{(n-3)}
\end{equation}

In Figure \@ref(fig:distcomparison)a we show an example distribution derived from the second expression (black curve), the inverse cumulative density function (red curve), and simulations from 100000 random data sets for n=10 (blue shading). The vertical lines show the critical (95%) threshold for the analytic and simulated results - a ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance. Figure \@ref(fig:distcomparison)b shows how these thresholds change as a function of the number of observations, and it is clear that the modified expression (red) most closely approximates the simulation results (blue).

```{r distcomparison, fig.cap='Panel (a) shows the distribution of condition indices derived from Equation 4 (black curve), and by simulation (blue shading), for a sample size of 10 observations. The vertical blue and red lines show the 95 percent thresholds on the curve (where 95 percent of values lie to the left of the line), and the red curve shows the inverse density function. Panel (b) shows how 95 percent thresholds change as a functio of the number of observations.', fig.align="center", fig.width=9, fig.height=5, echo=FALSE}

@@

plotlims <- c(1,5,0,1) 
ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2) 
title(xlab="Condition index", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)  
title(ylab="Probability", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)

@@

a <- hist(simvect,breaks = 200, plot=FALSE)
axvals <- a$mids
ayvals <- a$counts/max(a$counts)
polygon(c(1,axvals,1), c(0,ayvals,0), col=addalpha(nbdtpal[3],alpha=0.2),border=NA)
simcrit <- quantile(simvect,0.95)
lines(c(simcrit,simcrit),c(0,1),col=nbdtpal[3],lwd=2,lty=2)
text(4.8,0.95,'(a)',cex=1.5)

@@

lines(3:25,edelman1,lwd=2,col=nbdtpal[1])
lines(3:25,edelman2,lwd=2,col=nbdtpal[2])
lines(3:25,simratios,lwd=2,col=nbdtpal[3])
legend(12, 15, c("Edelman","Modified","Simulation"), cex=0.8, col=nbdtpal, lty=1, lwd=3, box.lwd=2)
text(23,24,'(b)',cex=1.5)

@@

A variant of the Mahalanobis distance (the pairwise Mahalanobis distance) can also be used to compute a multivariate measure of effect size, equivalent to Cohen's _d_ statistic [see e.g. @Giudice2009]. This is a valuable statistic to include when reporting the results of multivariate tests, and we include an _R_ function (_mahalES_) that can be used to calculate it.

# Generalising to more than two conditions
